\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Type-Preserving Mutations

### 2021-04-09 papers/oopsla-21/main.tex
Added abstract. Grammar and syntax check.

\begin{abstract}
  Nullam eu ante vel est convallis dignissim. Fusce suscipit, wisi nec facilisis
  facilisis, est dui fermentum leo, quis tempor ligula erat quis odio. Nunc
  porta vulputate tellus. Nunc rutrum turpis sed pede. Sed bibendum. Aliquam
  posuere. Nunc aliquet, augue nec adipiscing interdum, lacus tellus malesuada
  massa, quis varius mi purus non odio. Pellentesque condimentum, magna ut
  suscipit hendrerit, ipsum augue ornare nulla, non luctus diam neque sit amet
  urna. Curabitur vulputate vestibulum lorem. Fusce sagittis, libero non
  molestie mollis, magna orci ultrices dolor, at vulputate neque nulla lacinia
  eros. Sed id ligula quis est convallis tempor. Curabitur lacinia pulvinar
  nibh. Nam a sapien.
\end{abstract}

@@

Users of \quickcheck (and PBT in general) are well aware that one of the biggest
challenges while using these frameworks is to provide suitable random data
generators needed to instantiate the testing properties.
%
In some extreme cases, writing highly-tuned random generators capable of
exercising every part of a complex system in a reasonable basis can take several
thouthand person-hours of trial and error \cite{lampropoulos2019coverage}.

To alleviate this, there exist several techniques to automatically synthesize
random data generators by reyfing the static information present in the
codebase, e.g., data type definitions, application public interfaces (APIs),
etc. \cite{GriecoCB16, DBLP:conf/haskell/MistaRH18, Mista2019GeneratingRS,
  DuregardJW12, Lampropoulos2017, Bendkowski2017}.
%
These techniques, however, are unable to synthesize random generators capable of
producing data satisfying complex invariants not easily derivable from the
codebase.
%
Randomly generating valid programs is common example of this problem, where
developers are forced to put a substantial efforts in writing specialized random
generators by hand \cite{Palka11, perenyi2020stack, yang2011finding}.
%
Moreover, automatically synthesized random generators offer particularly poor
results when the testing properties are constrained by sparse preconditions.
%
In this scenario, most of the generated tests are simply discarded before they
can be used to test the main components of our system.

@@

Instead of continuously generating random invalid test cases from scratch,
\fuzzchick keeps a queue of interesting previously executed test cases that can
be mutated using type-preserving transformations in order to produce new ones.
%
Intuitively, mutating an interesting test case in a small way (at the data
constructor level) has a much higher probability of producing a new interesting
test case than generating a new one from scratch using a na\"ive generator.
%
In this light, \fuzzchick is likely to preserve the semantic structure of the
mutated data, as mutations are applied directly at the data type level, e.g.,
the program's AST in case of generating code.

@@

%% Problems of FuzzChick: mutations are superficial, long queues, power
%% scheduler
%
In this work, we take a close look at \fuzzchick for opportunities to improve.
%
Notably, we establish several aspects of its implementation that can be revised
--- some of these already being identified by the authors of \fuzzchick in their
original work.
%
In particular:
%
\begin{inparaenum}
\item \emph{the type-preserving mutations produced by this tool are superficial
  and rely heavily on randomness to produce diverse mutants},
\item \emph{using normal queues can cause large delays when interesting values
  are enqueued frequently}, and
\item \emph{the mechanism used to assing a certain mutation energy to each
  interesting test case requires fine tuning and can be hard to generalize to
  work well under different testing scenarios.}
\end{inparaenum}

@@

%% Exhaustive mutations
Unlike \fuzzchick, mutations in \mutagen are applied \emph{exhaustively}.
%
This is, given an interesting, previously executed test case, our tool
precomputes and schedules every type-presserving mutation that can be applied to
it.
%
This approach has two inherent advantages when compared to \fuzzchick.
%
On one hand, mutations do not rely on randomness to be generated.
%
Instead, every subexpression of the input test case is mutated on the same
basis, which guarantees that no interesting mutation is left behind due to
randomness.
%
On the other hand, scheduling mutations exaustively eliminates the need of a
heuristic that assigns a given mutation energy to each input test case.

@@

In this section, we briefly introduce the concept of Property-Based Testing
along with \quickcheck, one of the most popular tools of this sort and often
used as the baseline when comparing PBT algoritms.
%
Moreover, we describe in detail the ideas and limitations behind \fuzzchick,
which served as the foundation while designing \mutagen.

@@

\noindent For simplicity, here we instantiate the tree payload with integers,
altough this predicate should clearly hold for any other type with a properly
defined notion of equality as well.

@@

\noindent Let us break this definition into parts.
%
The first line states that we will provide an \texttt{Arbitrary} instance for
trees with payload of type \texttt{a}, provided that values of type \texttt{a}
can also be randomly generated.
%
This allows us to use \texttt{arbitrary} to generate \texttt{a}'s inside the
definition of our tree generator.

Moreover, \quickcheck internally keeps track of the \emph{maximum generation
  size}, a parameter that can be tuned by the user in order to limit the size of
the randomly generated values.
%
Our definition exposes this internal parameter via \quickcheck's \texttt{sized}
combinator, allowing us to parameterize the maximum size of the randomly
generated trees.
%
If the generation size is zero (\texttt{gen 0}), our generator is limited to
produce just leaves with randomly generated payloads.
%
In turn, when the generation size is strictly positive (\texttt{gen n}), the
generator is able to perform a random uniform choice between generating either a
single leaf or a branch.
%
When generating branches, the generator calls itself recursively in order to
produce random subtrees (\texttt{gen (n-1)}).
%
Notice the importance of reducing the generation size on each recursive call.
%
This way we ensure that randomly generated trees using a generation size
\texttt{n} are always finite and have at most \texttt{n} levels.

@@

In particular, \emph{automatically derived generators are remarkably uneffective
  when used to test properties with sparse preconditions.}
%
Let us continue with the example by \citeauthor{lampropoulos2019coverage} to
illustrate this problem in more detail.
%
For this, consider that we want to use our \texttt{Tree} data type to encode
binary-search trees (BST) --- this requires some minor tweaks in practice.
%
Then, given a predicate \texttt{isBST} that asserts if a tree satisfies the BST
invariants, we might want to use it as pre- and post-condition to assert that
BST operations like \texttt{insert} preserve them:

@@

At this point it is reasonable to think that, to obtain the best results when
using PBT over complex systems, one is forced to put a large amount of time on
developing manually-written generators.
%
In practice, that is most often the case, no automatic effort can beat a
well-thought manually-written generator that produces interesting complex values
and finds bugs in very few tests.
%
Not all is lost, however.
%
It is still possible to obtain acceptable results automatically by incorporating
dynamic information from the system under test into the testing loop.
%
The next subsection introduces the clever technique used by \fuzzchick to find
bugs in complex systems while using simple automatically derived random
generators.

@@

Since it is possible to distinguish semantically valid test cases from invalid
ones, i.e., those passing the sparse preconditions of our testing properties as
opossed to those that are discarded early, \fuzzchick exploits this information
in order to focus the testing efforts into mutating valid test cases with a
higher priority than those that were discarded.

@@

The energy assigned to each test case follows that of AFL's power schedule: more
energy to test cases that lead to shorter executions, or that discover more
parts of the code.
%
Moreover, to favour mutating interesting valid test cases, they get more energy
than those that were discarded.

@@

Mutators in \fuzzchick are no more than specialized random generators,
parameterized by the original input to be mutated.
%
They use a simple set of mutation operations that are randomly applied at the
data type level.
%
In simple terms, these operations encompass
%
\begin{inparaenum}
\item \emph{shrinking the value}, replacing its outtermost data constructor with
  one that contains a subset of its fields, reusing existing subexpressions;
\item \emph{growing the value}, replacing its outtermost data constructor with
  one that contains a superset of its fields, reusing existing subexpressions
  and generating random ones when needed;
\item \emph{returning a subexpression of the same type}; and
\item \emph{mutating recursively}, applying a mutation operation over an
  immediate subexpression.
\end{inparaenum}

\begin{figure}
\begin{verbatim}
mutate_tree :: (a -> Gen a) -> Tree a -> Gen (Tree a)
mutate_tree mutate_a (Leaf x) =
  oneof [ do { x' <- mutate_a x; return (Leaf x')}                       -- Mutate recursively
        , do { l <- arbitrary; r <- arbitrary; return (Branch l x r) } ] -- Grow constructor
mutate_tree mutate_a (Branch l x r) =
  oneof [ return l                                                       -- Return subexpression
        , return r                                                       -- Return subexpression
        , return (Leaf x)                                                -- Shrink constructor
        , do { l' <- mutate_tree l; return (Branch l' x r) }             -- Mutate recursively
        , do { x' <- mutate_a    x; return (Branch l x' r) }             -- Mutate recursively
        , do { r' <- mutate_tree r; return (Branch l x r') } ]           -- Mutate recursively
\end{verbatim}
\caption{\label{fig:fuzzchick:mutator}\fuzzchick mutator for the \texttt{Tree} data type. }
\end{figure}

@@

In this mutator, branches can be shrinked into leaves by dropping the subtrees,
whereas leaves can grow into branches, by reusing the payload and generating two
random subtrees.
%
Moreover, branches can be replaced with one of their subtrees.
%
Finally, mutations can be recursively applied over both the payload and the
subtrees.
%
At the top level, all these operations are put together using the \texttt{oneof}
combinator that randomly picks one of them with uniform probability.

@@

However, the authors acknowledge that certain parts of its implementation have
room for improvement, especially when it comes to the mutators design.
%
In tandem with the lack of efficacy observed when replicating the evaluation of
the IFC stack machine case study, these observations led us to consider three
main aspects of \fuzzchick that can be improved upon --- and that constitute the
main goal of this work.
%
%% As mentioned earlier, when we recreated the evaluation of the IFC stack machine
%% case study (described in detail in Section \ref{sec:casestudies}), we found that
%% after 30 runs (as opposed to the 5 runs used original by
%% \citeauthor{lampropoulos2019coverage}), \emph{\fuzzchick was only able to find 5
%%   out of the 20 injected bugs with a failure rate of 1}, the hardest one being
%% found only around 13\% of the time after an hour of testing.
%% %
%% These results are presented in detail in Section \ref{sec:evaluation}.
%
%% At the light of these observations, we identified three main aspects of
%% \fuzzchick that can be improved upon --- and that constitute the main goal of
%% this work.
%
In no particular order:

\paragraph{Mutators distribution:}
%
if we inspect the mutator defined in Fig. \ref{fig:fuzzchick:mutator}, there are
two compromises that the authors of \fuzzchick adopted for the sake of
simplicity.
%
On one hand, deep recursive mutations are very unlikely, since their probability
decreases multiplicatively with each recursive call.
%
For instance, mutating a subexpression that lies on the fourth level of a
\texttt{Tree} happens with a probability smaller than $(1/6)^3 = \sim 0.0046$,
i.e., when applying a recursive mutation (with probability $1/6$) three
consecutive times.
%
This only worsens as the type of the mutated value becomes more complex.
%
Hence, \fuzzchick mutators can only be effectively used to transform to shallow
data structures, potentially excluding interesting applications that might
require producing deeper valid values, e.g., programs, network protocols
interactions, etc.
%
Ideally, mutations should be able to happen on every subexpression of the input
seed in a reasonable basis.

On the other hand, using random generators to produce neeeded subexpressions
when growing data constructors can be dangerous, as we are introducing the very
same ``uncontrolled'' randomness that we wanted to mitigate in the first place!
%
If the random generator produces an invalid subexpression (something quite
likely), this might just invalidate the whole mutated test case.
%
We believe that growing data constructors needs to be done carefully.
%
For instance, by using just a minimal piece of data to make the overall mutated
test case type correct.
%
If that mutated test case turns to be interesting, that subexpression can always
be mutated later.

\paragraph{Enqueuing mutation candidates:}
%
\fuzzchick uses two single queues for keeping valid and discarded mutation
candidates.
%
Whenever a new test case is found interesting, it is placed \emph{at the end of
  its corresponding queue}.
%
If this test case happens to have discovered a whole new portion of the target
code, it will not be further mutated until the rest of the queue ahead of it
gets processed.
%
This can limit the effectiveness of the testing loop if the queues tend to grow
more often than they tend to shrink, as interesting mutation candidates can get
buried at the end of a long queue that only exercises the same portion of the
target code.
%
In the extreme case, they might not processed at all within the testing budget.
%
Ideally, one would like a mechanism that prioritizes mutating test cases that
discovers new portions of the code right away, and that is capable of jumping
back and forth from mutation candidates whenever this happens.
%
%% We show in Section \ref{sec:heuristics} how this can be achieved by analyzing
%% the execution information in order to prioritize test cases with novel execution
%% traces.

\paragraph{Power schedule:}
%
It is not clear how the power schedule used to assign energy to each mutable
test case in \fuzzchick works in the context of high-level type-preserving
mutations.
%
If it assigns too much energy to certain not-so-interesting seeds, some bugs
might not be discovered in a timely basis.
%
Conversely, assigning too little energy to interesting test cases might cause
that some bugs cannot be discovered at all unless the right mutation happens
within the small available energy window --- randomly generating the same test
case later on does not help, as it becomes uninteresting based on historic trace
information.

@@

In constrast with \fuzzchick, \mutagen does not employ a power schedule to
assign energy to mutable candidates.
%
In turn, it resorts to mutate them in an exhaustive and precise basis, where
%
\begin{inparaenum}
  \item each subexpression of a mutation candidate is associated with a set of
    type-preserving mutations, and
  \item for every mutable subexpression, each one of these mutations is
    evaluated \emph{exactly once}.
\end{inparaenum}
%
%% There is a small exception to this rules that we will introduce soon.

This idea is inspired by exhaustive bounded testing tools like SmallCheck
\cite{runciman2008smallcheck} or Korat \cite{boyapati2002korat}, that produce
test cases exhaustively.
%
In simple words, such tools work by enumerating all possible values of the
data types used in the testing properties, and then executing them from smaller
to larger until a a bug is found, or a certain size bound is reached.
%
The main problem with this approach is that the space of all possible test cases
often grows exponentially as we increment the size bound, and the user
experiences what it looks like ``hitting a wall'', where no larger test cases
can be evaluated until we exhausted all the immediately smaller ones
\cite{DuregardJW12}.
%
To alleviate this problem, such tools often rely on heuristics to prune the
search space based on detecting unevaluated subexpressions --- this is discussed
in Section \ref{sec:related}.

@@

Pure mutants are used most of the time, and encode simple deterministic
transformations over the outtermost data constructor of the input --- recursive
mutations will be introduced soon.
%
These tranformations can:
%
\begin{inparaenum}
\item return an immediate subexpression of the same type as the input;
\item swap the outtermost data constructor with any other constructor of the
  same type, reusing existing subexpressions; and
\item rearrange and replace fields using existing ones of the same type.
\end{inparaenum}
%
To illustrate this, Fig. \ref{fig:mutagen:mutator} outlines a mutator for the
\texttt{Tree} data type.
%
Notice how this definition simply enumerates mutants that transform the
outtermost data constructor, hence no recursion is needed here.
%
Moreover, notice how a default value (\texttt{def}) used to fill the subtrees
when transforming a leaf into a branch.
%
This value corresponds to the simplest value we can construct in order for the
mutant to be type-correct.
%
In practice (\texttt{def = Leaf def}), where the inner \texttt{def} is the
simplest value of the payload --- we use the type class system to abstract this
complexity away in our implementation.
%
Using a small value is again inspired by exhaustive bounded testing tools, and
avoids introducing unnecesary randomness when growing data constructors.

\begin{figure}
\begin{verbatim}
mutate (Leaf x)       = [ PURE (Branch def x def) ]   -- Swap constructor
mutate (Branch l x r) = [ PURE l, PURE r              -- Return subexpression
                        , PURE (Leaf x)               -- Swap constructor
                        , PURE (Branch l x l)         -- Rearrange subexpressions
                        , PURE (Branch r x r)         -- Rearrange subexpressions
                        , PURE (Branch r x l) ]       -- Rearrange subexpressions
\end{verbatim}
\caption{\label{fig:mutagen:mutator}\mutagen mutator for the \texttt{Tree} data
  type.}
\end{figure}

%% At this point, the reader might consider:
%% %
%% what about large enumeration types like integers or characters?
%% %
%% Does \mutagen transform them exaustively too?
%% %
%% \emph{Certainly not.}
%
Random mutants, on the other hand, serve as a way to break exhaustiveness when
mutating values of large enumeration types --- or any other type the user might
want to introduce random mutations to.
%
Instead of mutating into every other possible number or character, we resolve in
using a random generator and sample a small amount of values from it --- the
precise amount is a tunable parameter of \mutagen.
%
This way, a mutator for integers simply becomes:

@@

\noindent where the function \texttt{node} simply builds a node of the positions
rose tree and propagates the accummulated prefix (the path from the root) to its
children.
%
In this light, the mutable positions of the value \texttt{Branch (Leaf 1) 2
  (Leaf 3)} are:
%
\begin{equation}
  \nonumber
  {\small \texttt{positions}}\quad
  \left(
  \tikz[
    baseline=-7mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{Branch}}
      child {node {\small \texttt{Leaf}}
        child {node {\small \texttt{1}}}
      }
      child {node {\small \texttt{2}}}
      child {node {\small \texttt{Leaf}}
        child {node {\small \texttt{3}}}
      };
  }
  \right)
  \quad
  =
  \quad
  \tikz[
    baseline=-7mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\texttt{\small []}}
      child {node {\small \texttt{[0]}}
        child {node {\small \texttt{[0,0]}}}
      }
      child {node {\small \texttt{[1]}}}
      child {node {\small \texttt{[2]}}
        child {node {\small \texttt{[2,0]}}}
      };
  }
\end{equation}

>>>>

\begin{abstract}
Automatically synthesized random data generators are an appealing option when
using property-based testing to validate our systems.
%
Such random generators can be obtained using a variety of techniques that work
by extracting static information from the codebase in order to produce random
test cases.
%
For the most part, however, such techniques are not suitable for deriving
generators producing random values satisfying complex invariants, which in turn
complicates testing properties with sparse preconditions.

\fuzzchick is a recent tool that borrows ideas from the fuzzing community to
alleviate this limitation.
%
It enhances automatically synthesized generators with type-preserving mutations
guided by execution traces.
%
Despite being enlighthening, we observed several limitations in the
implementation of \fuzzchick that reduce its applicability.
%
These limitations include a large reliance on randomness when mutating test
cases, as well as poor scheduling in certain situations, among others.

In this work we present \mutagen, a coverage-guided, property-based testing
framework that builds upon the ideas behind \fuzzchick.
%
Instead of relying heavily on randomness, \mutagen uses an exhaustive approach
to generate new mutants.
%
This is coupled with heuristics that help to schedule mutants based on their
novelty and to minimize mutating subexpressions of large enumeration types more
than necessary.
%
Our results indicate that our approach is capable of outperforming the simpler
one used by \fuzzchick, as well as finding subtle bugs in real-world scenarios.
\end{abstract}

@@

Users of \quickcheck (and PBT in general) are well aware that one of the biggest
challenges while using these frameworks is to provide suitable random data
generators needed to instantiate the testing properties.
%
In some extreme cases, writing highly-tuned random generators capable of
exercising every part of a complex system on a reasonable basis can take several
thousand person-hours of trial and error \cite{lampropoulos2019coverage}.

To alleviate this, there exist several techniques to automatically synthesize
random data generators by reifying the static information present in the
codebase, e.g., data type definitions, application public interfaces (APIs),
etc. \cite{GriecoCB16, DBLP:conf/haskell/MistaRH18, Mista2019GeneratingRS,
  DuregardJW12, Lampropoulos2017, Bendkowski2017}.
%
These techniques, however, are unable to synthesize random generators capable of
producing data satisfying complex invariants not easily derivable from the
codebase.
%
Randomly generating valid programs is a common example of this problem, where
developers are forced to put substantial efforts in writing specialized random
generators by hand \cite{Palka11, perenyi2020stack, yang2011finding}.
%
Moreover, automatically synthesized random generators offer particularly poor
results when the testing properties are constrained by sparse preconditions.
%
In this scenario, most of the generated tests are simply discarded before they
can be used to test the main components of our system.

@@

Instead of continuously generating random invalid test cases from scratch,
\fuzzchick keeps a queue of interesting previously executed test cases that can
be mutated using type-preserving transformations to produce new ones.
%
Intuitively, mutating an interesting test case in a small way (at the data
constructor level) has a much higher probability of producing a new interesting
test case than generating a new one from scratch using a na\"ive generator.
%
In this light, \fuzzchick is likely to preserve the semantic structure of the
mutated data, as mutations are applied directly at the data type level, e.g.,
the program's AST in case of generating code.

@@

%% Problems of FuzzChick: mutations are superficial, long queues, power
%% scheduler
%
In this work, we take a close look at \fuzzchick for opportunities to improve.
%
Notably, we establish several aspects of its implementation that can be revised
--- some of these already being identified by the authors of \fuzzchick in their
original work.
%
In particular:
%
\begin{inparaenum}
\item \emph{the type-preserving mutations produced by this tool are superficial
  and rely heavily on randomness to produce diverse mutants},
\item \emph{using normal queues can cause large delays when interesting values
  are enqueued frequently}, and
\item \emph{the mechanism used to assign a certain mutation energy to each
  interesting test case requires fine tuning and can be hard to generalize to
  work well under different testing scenarios.}
\end{inparaenum}

@@

%% Exhaustive mutations
Unlike \fuzzchick, mutations in \mutagen are applied \emph{exhaustively}.
%
This is, given an interesting, previously executed test case, our tool
precomputes and schedules every type-preserving mutation that can be applied to
it.
%
This approach has two inherent advantages when compared to \fuzzchick.
%
On one hand, mutations do not rely on randomness to be generated.
%
Instead, every subexpression of the input test case is mutated on the same
basis, which guarantees that no interesting mutation is left behind due to
randomness.
%
On the other hand, scheduling mutations exhaustively eliminates the need of a
heuristic that assigns a given mutation energy to each input test case.

@@

In this section, we briefly introduce the concept of Property-Based Testing
along with \quickcheck, one of the most popular tools of this sort and often
used as the baseline when comparing PBT algorithms.
%
Moreover, we describe in detail the ideas and limitations behind \fuzzchick,
which served as the foundation while designing \mutagen.

@@

\noindent For simplicity, here we instantiate the tree payload with integers,
although this predicate should hold for any other type with a properly defined
notion of equality as well.

@@

\noindent Let us break this definition into parts.
%
The first line states that we will provide an \texttt{Arbitrary} instance for
trees with a payload of type \texttt{a}, provided that values of type \texttt{a}
can also be randomly generated.
%
This allows us to use \texttt{arbitrary} to generate \texttt{a}'s inside the
definition of our tree generator.

Moreover, \quickcheck internally keeps track of the \emph{maximum generation
  size}, a parameter that can be tuned by the user to limit the size of the
randomly generated values.
%
Our definition exposes this internal parameter via \quickcheck's \texttt{sized}
combinator, allowing us to parameterize the maximum size of the randomly
generated trees.
%
If the generation size is zero (\texttt{gen 0}), our generator is limited to
produce just leaves with randomly generated payloads.
%
In turn, when the generation size is strictly positive (\texttt{gen n}), the
generator can perform a random uniform choice between generating either a single
leaf or a branch.
%
When generating branches, the generator calls itself recursively in order to
produce random subtrees (\texttt{gen (n-1)}).
%
Notice the importance of reducing the generation size on each recursive call.
%
This way we ensure that randomly generated trees using a generation size
\texttt{n} are always finite and have at most \texttt{n} levels.

@@

In particular, \emph{automatically derived generators are remarkably ineffective
  when used to test properties with sparse preconditions.}
%
Let us continue with the example by \citeauthor{lampropoulos2019coverage} to
illustrate this problem in more detail.
%
For this, consider that we want to use our \texttt{Tree} data type to encode
binary-search trees (BST) --- this requires some minor tweaks in practice.
%
Then, given a predicate \texttt{isBST} that asserts if a tree satisfies the BST
invariants, we might want to use it as pre- and post-condition to assert that
BST operations like \texttt{insert} preserve them:

@@

At this point it is reasonable to think that, to obtain the best results when
using PBT over complex systems, one is forced to put a large amount of time into
developing manually-written generators.
%
In practice, that is most often the case, no automatic effort can beat a
well-thought manually-written generator that produces interesting complex values
and finds bugs in very few tests.
%
Not all is lost, however.
%
It is still possible to obtain acceptable results automatically by incorporating
dynamic information from the system under test into the testing loop.
%
The next subsection introduces the clever technique used by \fuzzchick to find
bugs in complex systems while using simple automatically derived random
generators.

@@

Since it is possible to distinguish semantically valid test cases from invalid
ones, i.e., those passing the sparse preconditions of our testing properties as
opposed to those that are discarded early, \fuzzchick exploits this information
in order to focus the testing efforts into mutating valid test cases with a
higher priority than those that were discarded.

@@

The energy assigned to each test case follows that of AFL's power schedule: more
energy to test cases that lead to shorter executions, or that discover more
parts of the code.
%
Moreover, to favor mutating interesting valid test cases, they get more energy
than those that were discarded.

@@

Mutators in \fuzzchick are no more than specialized random generators,
parameterized by the original input to be mutated.
%
They use a simple set of mutation operations that are randomly applied at the
data type level.
%
In simple terms, these operations encompass
%
\begin{inparaenum}
\item \emph{shrinking the value}, replacing its outermost data constructor with
  one that contains a subset of its fields, reusing existing subexpressions;
\item \emph{growing the value}, replacing its outermost data constructor with
  one that contains a superset of its fields, reusing existing subexpressions
  and generating random ones when needed;
\item \emph{returning a subexpression of the same type}; and
\item \emph{mutating recursively}, applying a mutation operation over an
  immediate subexpression.
\end{inparaenum}

\begin{figure}
\begin{verbatim}
mutate_tree :: (a -> Gen a) -> Tree a -> Gen (Tree a)
mutate_tree mutate_a (Leaf x) =
  oneof [ do { x' <- mutate_a x; return (Leaf x')}                       -- Mutate recursively
        , do { l <- arbitrary; r <- arbitrary; return (Branch l x r) } ] -- Grow constructor
mutate_tree mutate_a (Branch l x r) =
  oneof [ return l                                                       -- Return subexpression
        , return r                                                       -- Return subexpression
        , return (Leaf x)                                                -- Shrink constructor
        , do { l' <- mutate_tree l; return (Branch l' x r) }             -- Mutate recursively
        , do { x' <- mutate_a    x; return (Branch l x' r) }             -- Mutate recursively
        , do { r' <- mutate_tree r; return (Branch l x r') } ]           -- Mutate recursively
\end{verbatim}
\vspace{-5pt}
\caption{\label{fig:fuzzchick:mutator}\fuzzchick mutator for the \texttt{Tree} data type. }
\vspace{-10pt}
\end{figure}

@@

In this mutator, branches can be shrunk into leaves by dropping the subtrees,
whereas leaves can grow into branches, by reusing the payload and generating two
random subtrees.
%
Moreover, branches can be replaced with one of their subtrees.
%
Finally, mutations can be recursively applied over both the payload and the
subtrees.
%
At the top level, all these operations are put together using the \texttt{oneof}
combinator that randomly picks one of them with uniform probability.

@@

However, the authors acknowledge that certain parts of its implementation have
room for improvement, especially when it comes to the mutator's design.
%
In tandem with the lack of efficacy observed when replicating the evaluation of
the IFC stack machine case study, these observations led us to consider three
main aspects of \fuzzchick that can be improved upon --- and that constitute the
main goal of this work.
%
%% As mentioned earlier, when we recreated the evaluation of the IFC stack machine
%% case study (described in detail in Section \ref{sec:casestudies}), we found that
%% after 30 runs (as opposed to the 5 runs used original by
%% \citeauthor{lampropoulos2019coverage}), \emph{\fuzzchick was only able to find 5
%%   out of the 20 injected bugs with a failure rate of 1}, the hardest one being
%% found only around 13\% of the time after an hour of testing.
%% %
%% These results are presented in detail in Section \ref{sec:evaluation}.
%
%% At the light of these observations, we identified three main aspects of
%% \fuzzchick that can be improved upon --- and that constitute the main goal of
%% this work.
%
In no particular order:

\paragraph{Mutators distribution:}
%
if we inspect the mutator defined in Fig. \ref{fig:fuzzchick:mutator}, there are
two compromises that the authors of \fuzzchick adopted for the sake of
simplicity.
%
On one hand, deep recursive mutations are very unlikely, since their probability
decreases multiplicatively with each recursive call.
%
For instance, mutating a subexpression that lies on the fourth level of a
\texttt{Tree} happens with a probability smaller than $(1/6)^3 = \sim 0.0046$,
i.e., when applying a recursive mutation (with probability $1/6$) three
consecutive times.
%
This only worsens as the type of the mutated value becomes more complex.
%
Hence, \fuzzchick mutators can only be effectively used to transform to shallow
data structures, potentially excluding interesting applications that might
require producing deeper valid values, e.g., programs, network protocol
interactions, etc.
%
Ideally, mutations should be able to happen on every subexpression of the input
seed on a reasonable basis.

On the other hand, using random generators to produce needed subexpressions
when growing data constructors can be dangerous, as we are introducing the very
same ``uncontrolled'' randomness that we wanted to mitigate in the first place!
%
If the random generator produces an invalid subexpression (something quite
likely), this might just invalidate the whole mutated test case.
%
We believe that growing data constructors needs to be done carefully.
%
For instance, by using just a minimal piece of data to make the overall mutated
test case type correct.
%
If that mutated test case turns to be interesting, that subexpression can always
be mutated later.

\paragraph{Enqueuing mutation candidates:}
%
\fuzzchick uses two single queues for keeping valid and discarded mutation
candidates.
%
Whenever a new test case is found interesting, it is placed \emph{at the end of
  its corresponding queue}.
%
If this test case happens to have discovered a whole new portion of the target
code, it will not be further mutated until the rest of the queue ahead of it
gets processed.
%
This can limit the effectiveness of the testing loop if the queues tend to grow
more often than they tend to shrink, as interesting mutation candidates can get
buried at the end of a long queue that only exercises the same portion of the
target code.
%
In an extreme case, they might not be processed at all within the testing
budget.
%
Ideally, one would like a mechanism that prioritizes mutating test cases that
discover new portions of the code right away, and that is capable of jumping
back and forth from mutation candidates whenever this happens.
%
%% We show in Section \ref{sec:heuristics} how this can be achieved by analyzing
%% the execution information in order to prioritize test cases with novel execution
%% traces.

\paragraph{Power schedule:}
%
It is not clear how the power schedule used to assign energy to each mutable
test case in \fuzzchick works in the context of high-level type-preserving
mutations.
%
If it assigns too much energy to certain not-so-interesting seeds, some bugs
might not be discovered on a timely basis.
%
Conversely, assigning too little energy to interesting test cases might cause
that some bugs cannot be discovered at all unless the right mutation happens
within the small available energy window --- randomly generating the same test
case again later does not help, as it becomes uninteresting based on historic
trace information.

@@

In contrast with \fuzzchick, \mutagen does not employ a power schedule to assign
energy to mutable candidates.
%
In turn, it resorts to mutate them on an exhaustive and precise manner, where
%
\begin{inparaenum}
  \item each subexpression of a mutation candidate is associated with a set of
    type-preserving mutations, and
  \item for every mutable subexpression, each one of these mutations is
    evaluated \emph{exactly once}.
\end{inparaenum}
%
%% There is a small exception to this rules that we will introduce soon.

This idea is inspired by exhaustive bounded testing tools like SmallCheck
\cite{runciman2008smallcheck} or Korat \cite{boyapati2002korat}, that produce
test cases exhaustively.
%
In simple words, such tools work by enumerating all possible values of the data
types used in the testing properties, and then executing them from smaller to
larger until a bug is found, or a certain size bound is reached.
%
The main problem with this approach is that the space of all possible test cases
often grows exponentially as we increment the size bound, and the user
experiences what it looks like ``hitting a wall'', where no larger test cases
can be evaluated until we exhausted all the immediately smaller ones
\cite{DuregardJW12}.
%
To alleviate this problem, such tools often rely on heuristics to prune the
search space based on detecting unevaluated subexpressions --- this is discussed
in Section \ref{sec:related}.

@@

Pure mutants are used most of the time, and encode simple deterministic
transformations over the outermost data constructor of the input --- recursive
mutations will be introduced soon.
%
These transformations can:
%
\begin{inparaenum}
\item return an immediate subexpression of the same type as the input;
\item swap the outermost data constructor with any other constructor of the
  same type, reusing existing subexpressions; and
\item rearrange and replace fields using existing ones of the same type.
\end{inparaenum}
%
To illustrate this, Fig. \ref{fig:mutagen:mutator} outlines a mutator for the
\texttt{Tree} data type.
%
Notice how this definition simply enumerates mutants that transform the
outermost data constructor, hence no recursion is needed here.
%
Moreover, notice how a default value (\texttt{def}) used to fill the subtrees
when transforming a leaf into a branch.
%
This value corresponds to the simplest expression we can construct for the
mutant to be type-correct.
%
In practice (\texttt{def = Leaf def}), where the inner \texttt{def} is the
simplest value of the payload --- we use the type class system to abstract this
complexity away in our implementation.
%
Using a small value is again inspired by exhaustive bounded testing tools, and
avoids introducing unnecessary randomness when growing data constructors.

\begin{figure}
\begin{verbatim}
mutate (Leaf x)       = [ PURE (Branch def x def) ]   -- Swap constructor
mutate (Branch l x r) = [ PURE l, PURE r              -- Return subexpression
                        , PURE (Leaf x)               -- Swap constructor
                        , PURE (Branch l x l)         -- Rearrange subexpressions
                        , PURE (Branch r x r)         -- Rearrange subexpressions
                        , PURE (Branch r x l) ]       -- Rearrange subexpressions
\end{verbatim}
\vspace{-5pt}
\caption{\label{fig:mutagen:mutator}\mutagen mutator for the \texttt{Tree} data
  type.}
\vspace{-10pt}
\end{figure}

%% At this point, the reader might consider:
%% %
%% what about large enumeration types like integers or characters?
%% %
%% Does \mutagen transform them exaustively too?
%% %
%% \emph{Certainly not.}
%
Random mutants, on the other hand, serve as a way to break exhaustiveness when
mutating values of large enumeration types --- or any other type the user might
want to use random mutations with.
%
Instead of mutating into every other possible number or character, we resolve in
using a random generator and sample a small number of values from it --- the
precise amount is a tunable parameter of \mutagen.
%
This way, a mutator for integers simply becomes:

@@

\noindent where the function \texttt{node} simply builds a node of the positions
rose tree and propagates the accumulated prefix (the path from the root) to its
children.
%
In this light, the mutable positions of the value \texttt{Branch (Leaf 1) 2
  (Leaf 3)} are:
%
\begin{equation}
  \nonumber
  {\small \texttt{positions}}\quad
  \left(
  \tikz[
    baseline=-7mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{Branch}}
      child {node {\small \texttt{Leaf}}
        child {node {\small \texttt{1}}}
      }
      child {node {\small \texttt{2}}}
      child {node {\small \texttt{Leaf}}
        child {node {\small \texttt{3}}}
      };
  }
  \right)
  \quad
  =
  \quad
  \tikz[
    baseline=-7mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\texttt{\small []}}
      child {node {\small \texttt{[0]}}
        child {node {\small \texttt{[0,0]}}}
      }
      child {node {\small \texttt{[1]}}}
      child {node {\small \texttt{[2]}}
        child {node {\small \texttt{[2,0]}}}
      };
  }
\end{equation}

\noindent Later, given the desired position to mutate within a test case, we
define another function \texttt{inside} that finds the subexpression
corresponding to it and applies the mutation.
%
A slightly simplified version of this function for the \texttt{Tree} data type is
as follows:
